### 2013-09-29 manuscripts/dpp-msbayes/dpp-msbayes.tex
Finishing initial draft of dpp ms.

        \vspace{12pt}
        \textbf{Running head:} \MakeUppercase{Running head}

@@

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent Abstract here \ldots

    \vspace{12pt}
    \noindent\textbf{KEY WORDS: } 
}

\newpage
\noindent \ldots

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of Evolutionary Biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can effect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history,
e.g., temporally clustered divergences, across effected populations of
organisms.
At recent timescales, such events should leave a signature in the genetic
variation within and among the effected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\citep{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and coalescence.

Such a method has been developed in an approximate-Bayesian model-choice
framework and is implemented in the software package \msb
\citep{Hickerson2006,Huang2011}.
Recently, however, \citet{Oaks2012} found the method can be strongly biased
toward spuriously estimating biogeographically interesting models of highly
clustered divergence times across taxa.
The results of their simulation-based assessment of the method suggest that a
combination of broad, uniformly distributed priors on nuisance parameters and a
U-shaped prior on divergence models were causing the bias.
The model implemented in \msb uses a discrete uniform prior over the number of
possible divergence events, which \citet{Oaks2012} show imposes a prior on
divergence models that puts most of the prior mass on models with either very
few or very many divergence times shared across the taxa.
Furthermore, the uniform priors on most of the model's nuisance parameters
likely hinder the marginal likelihoods of parameter-rich models, i.e., models
with many divergence time parameters \citep[this is often referred to as
Lindley's paradox;][]{Lindley1957}.
Thus, together, the priors on divergence models and nuisance parameters used in
\msb could create a strong bias toward models with very few divergence time
parameters.

Following the recommendations of \citet{Oaks2012} we modify the \msb model to
test whether alternative parameterizations and priors will improve the behavior
of the approximate Bayesian model-choice approach.
We implement a Dirichlet-process prior over all possible models of divergence,
and use gamma and beta probability distributions in place of uniform priors on
many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results show that both the uniform priors on nuisance parameters and
the U-shaped prior on divergence model contribute to the biases of \msb.

>>>>

        \vspace{12pt}
        \textbf{Running head:} \MakeUppercase{Improved Estimation of Shared Evolutionary History}

@@

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that effect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated from genetic data from co-distributed species.
    We introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    Our model is an extension of the model implemented in \msb.
    By reparameterizing the model, and using more flexible priors on divergence
    models and nuisance parameters, we improve the robustness, accuracy, and power
    of the method for estimating the posterior probabilities of models
    of divergence across taxa.
    Our results demonstrate that the bias of the \msb model toward inferring
    models of clustered divergences is caused by a combination of (1) uniform
    priors on nuisance parameters hindering the marginal likelihoods of models
    with more divergence time parameters, and (2) a prior on divergence models
    that disfavors models with intermediate numbers of divergence time
    parameters.

    \vspace{12pt}
    \noindent\textbf{KEY WORDS: Dirichlet-process prior; approximate Bayesian
    computation; model choice; phylogeography; biogeography} 
}

\newpage

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of Evolutionary Biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can effect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across effected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the effected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\citep{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and lineage coalescence.

The software package \msb implements such an approach in an
approximate-Bayesian model-choice framework \citep{Hickerson2006,Huang2011}.
Recently, however, \citet{Oaks2012} found that \msb can be strongly biased
toward spuriously estimating biogeographically interesting models of highly
clustered divergence times across taxa.
The results of their simulation-based assessment of the method suggest that a
combination of broad, uniformly distributed priors on nuisance parameters and a
U-shaped prior on divergence models were causing the bias.
The model implemented in \msb uses a discrete uniform prior over the number of
possible divergence parameters, which \citet{Oaks2012} show imposes a prior on
divergence models that puts most of the prior mass on models with either very
few or very many divergence times shared across the taxa.
Furthermore, uniform priors used for most of the model's nuisance parameters
likely hinder the marginal likelihoods of parameter-rich models, i.e., models
with many divergence time parameters \citep[this is often referred to as
Lindley's paradox;][]{Lindley1957}.
Thus, together, the priors on divergence models and nuisance parameters used in
\msb could create a strong bias toward models with very few divergence time
parameters \citep{Oaks2012}.

Following the recommendations of \citet{Oaks2012} we modify the \msb model to
test whether alternative parameterizations and priors will improve the behavior
of the approximate Bayesian model-choice approach.
We implement a Dirichlet-process prior over all possible models of divergence,
and use gamma and beta probability distributions in place of uniform priors on
many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results show that both the uniform priors on nuisance parameters and
the U-shaped prior on divergence models contribute to the biases of \msb.
